An antibody-escape calculator for mutations to the SARS-CoV-2 receptor-binding domain

### 2021-12-04 paper/paper.tex
start making Omicron figures
added Omicron results subsection

\subsection{The escape calculations correlate well with neutralization assays on SARS-CoV-2 variants}
We can also define a quantitative score that summarizes the polyclonal antibody activity that remains after an arbitrary combination of RBD mutations.
This score is defined using the same principle as the site-wise escape calculator in the previous subsection: we reduce the contribution of each antibody by an amount that scales with how strongly it is escaped by each mutated site, and define the overall score as the fraction of all antibody contributions that remain after the mutations (see Methods for details).
This calculation is implemented in the interactive calculator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, and returns a score that ranges from one (no mutations affect binding of any antibodies) to zero (all antibodies fully escaped).

@@

\subsection{}
Omicron is 0.2.
PMS20 is 0.19

>>>>

\subsection{The escape calculations correlate well with neutralization assays on SARS-CoV-2 variants}
We can also define a quantitative score that summarizes the polyclonal antibody binding that remains after an arbitrary combination of RBD mutations.
This score is defined using the same principle as the site-wise escape calculator in the previous subsection: we reduce the contribution of each antibody by an amount that scales with how strongly it is escaped by each mutated site, and define the overall score as the fraction of all antibody contributions that remain after the mutations (see Methods for details).
This calculation is implemented in the interactive calculator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, and returns a score that ranges from one (no mutations affect binding of any antibodies) to zero (all antibodies fully escaped).

@@

\subsection{The escape calculator suggests extensive antigenic change in the new Omicron variant}
We applied the escape calculator to the recently reported Omicron variant, which has 15 mutated sites in its RBD~\citep{?}.
The overall escape-calculator binding score for the polyclonal antibody mix is much lower for the Omicron variant than any other SARS-CoV-2 variants of concern (Figure~\ref{fig:Omicron}A).
The Omicron variant's calculated score is roughly equivalent to that of a polymutant spike (PMS20) that was artificially engineered in a pseudovirus by \citet{schmidt2021high} to maximize escape from polyclonal serum antibodies.
For comparison, \citet{schmidt2021high} measured that neutralization titers against this artificial PMS20 spike were reduced by $\sim$20- to $\sim$80-fold for sera from various cohorts of vaccinated and infected individuals.

\begin{figure*}
\includegraphics[width=\linewidth]{figures/Omicron/Omicron.pdf}
\caption{Escape calculations for the Omicron variant.
{\bf (A)} The calculated binding scores for SARS-CoV-2 variants and the artificial polymutant spike (PMS20) generated by \citet{schmidt2021high}.
Scores of one indicate no mutations affect binding, and scores of zero indicate no antibody binding remains.
{\bf (B)} The calculated escape map for the Omicron variant's RBD (blue) compared to an unmutated RBD (gray), with sites of mutations in the Omicron variant in red.
}
\label{fig:Omicron}
\end{figure*}

The site-level escape map for the Omicron variant's RBD is shown in Figure~\ref{fig:Omicron}B.
That map shows that the Omicron RBD has lost most peaks of antibody binding relative to the original RBD.
Exploration of the mutations using the interactive calculator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/} indicates that the mutations at sites 417, 446, and 484 are the biggest drivers of this antigenic change, although other mutations also contribute.
The residual peaks in the map suggests that the remaining antibody activity against the Omicron variant RBD could be further eroded by mutations at sites like 346, 378, 444, and 504.
